How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-07-20 2.01-Hyp.Rmd
Add files via upload
# After copy edits

The writing part of the PhD is often the most difficult for many students. I think that there are several reasons for this. Writing well comes with practice, and most PhD students have had little experience with academic writing - even if they have already completed a MSc. The same could be said of most early career scientists (even after their PhDs), because practice is measured in completed pieces, it usually takes many years rather than weeks or months. YOu shouldn't expect that scientific writing will come automatically or immediately. In order to get this practice you need to get started, and (again) many students find starting to write challenging. This guide was written to help you get started.

@@

In this part of the book, I start with an explanation of the hypothesis. As you will see in [Part III](hypothesis.html), the hypothesis is central to all of the writing in each chapter or paper, and you need to have a very good idea of how to formulate your research question into a hypothesis before you begin. Then I build on helping you think through lots of different aspects of writing in science: style, citations, writing concisely, scientific names, and lots more. Finally, I introduce you to the idea of using a formula to structure your writing.

@@

When you read about the formula of writing a chapter (see end of part 2) you will see that everything in your chapter or paper revolves around the hypothesis or questions that you ask, normally at the end of your introduction. In this chapter, we will take a look at how to write a hypothesis. This is a sticking point for many students. We are used to using and writing questions and statements in day to day communications, as well as reading popular media. But hypotheses (the plural of hypothesis) only rarely float across our desks. So how do we write one, and how do we know if our hypothesis is good?

In addition to this chapter, there is some good information out there on the web, and it's worth looking at this too: (e.g. [Wikihow](https://www.wikihow.com/Write-a-Hypothesis), [Wikipedia](https://en.wikipedia.org/wiki/Hypothesis), etc.). There's also some less good stuff out there, so read critically.

(ref:Hypothesis) **Generating a good hypothesis for your PhD chapter isn't easy.** A good hypothesis will be invaluable in helping you write the chapter. It is important to start working with hypotheses

@@

A hypothesis is a statement of your research intent. It tells the reader (because just like all of your other written work, it is intended for an audience who will read it), what you plan to do in your research. But there's a little more to it than this. The hypothesis becomes a part of the scientific method if it is testable and (importantly) falsifiable, as well as being informed from previous published work on the subject.

Yes, your hypothesis must be informed by the literature, which is why you spent so much time and effort crafting your introduction to inform your reader of the same. This is also why your hypothesis usually comes at the end of your introduction, because you spend all of the introduction telling your reader about it. There's not much point in writing more after the hypothesis, because once your reader has read that, they are ready to learn about how you went about testing it (in the Materials & Methods). The other important point to make is that the literature should dictate how you write your hypothesis, and the variables that you include. If, for example, you know that temperature is the most important variable, but all of the literature suggests that it is oxygen, you can't ignore oxygen and you should also frame your hypothesis using this variable (you can have more than one hypothesis after all!). In this case, you will also need to provide a sufficient introduction to temperature as a variable to justify its inclusion in your hypothesis. Perversely, your aim is not to prove that your idea is right, but to show that the hypothesis can be refuted.

We try to write a hypothesis that is falsifiable: i.e. you can prove (usually using statistical tests) that it is not correct (or at least show that the likelihood that it is correct is very low). That's why it is conventional to provide the 'null hypothesis' that is the falsified version of the statement, suggesting that there is no relationship between the variables you have proposed to measure. The convention is to label this 'null hypothesis' H0, while the 'alternative hypothesis' (the one that says your variables are related as you suggested) is written as H1. When you formulate your hypothesis, it is traditional to write your alternative hypothesis to indicate the directionality of your tested variables. This way the reader can simply imply that the null hypothesis is when there is a relationship, but this will need to be stated if the null hypothesis is more complex.

@@

Most importantly, your hypothesis must come first, before you do the experiment or study. Hence the reason why this chapter comes at the start of part 2. Setting the hypothesis after the work is already done is fraudulent, and goes against the scientific method. Obviously, it isn't fair to pose the hypothesis once you already know the answer. This is why there is so much emphasis put on formulating your hypothesis during your research proposal. Getting it right will determine what you do and how you test it. If you think of an extra hypothesis that would be really useful to test once you've already done your study, you can conduct a post hoc test, but this should have more stringent levels of statistical assessment.

Writing a hypothesis isn't easy, but it is essential and once you've understood what to do most of the rest of what you are writing for should make sense.

@@

Later in the book, I will emphasise that you must have introduced all the variables that you plan to use to test your hypothesis in your introduction. This usually comes in the second paragraph of your introduction, where you emphasise the utility of the dependent variable/s (what you are planning to measure) and your independent variable (what you will manipulate). Both of these variables should then feature in your hypothesis. Next, by paragraph four you will have identified the problem that you are interested in tackling. In addition, your introduction will provide all of the pertinent literature that has relevance to this hypothesis, giving the all important context.

A simple way to consider making your hypothesis is to adopt an **"If. then. because."** construction where you add in your problem statement using your independent variable after 'if', and your prediction using your dependent variable after 'then', and finally the expected mechanism after 'because'. Using our example above with the "If. then. because." construction, we would say: "If environmental temperatures in which tadpoles develop are increased then tadpole development rate is faster because they follow the classic metabolism of ectotherms". Both independent variable (temperature) and dependent variable (tadpole development rate) are present in this hypothesis, and the predicted relationship between them is clear. In addition, the causal mechanism is stated. You can watch a video about using the "If. then. because." construction here, or read more here. I say that this is a formulaic way to start writing your hypothesis, because it usually ends up as an inelegant statement, which can be better refined for a reader. A citation for your stated mechanism might also help clarify exactly where the justification for this comes from.

@@

Exogenous causal explanations concern mechanisms that are outside the body of individuals. Common exogenous mechanisms are climatic factors (temperature, humidity, precipitation, etc.), or may relate to the availability of food, predators or mates.

@@

### Teleological versus Causal Hypotheses

A teleological argument refers to the reason or a purpose of a particular process. For example, you may measure vertical migration of water fleas and suggest that diurnal migrations are made because the water fleas want to avoid predation. This is a teleological hypothesis because you are suggesting that the reason behind a process is the desire by water fleas to avoid predators. Although a reduction of predation may be a consequence of vertical migration in water fleas, each water flea does not think about predation and then starts it’s upward movement as a result. A common mistake made in biology is to apply teleological arguments to processes that have no purpose or region. Evolution is often mistakenly suggested to have a purpose (e.g. to evolve to a more advanced state), but in fact evolution is not a goal-orientated process. There is no end-point to evolution, and evolution did not start in order to meet some predetermined form or function. On the other hand, a causal hypothesis focuses on the factors about A that cause B.

>>>>

The writing part of the PhD is often the most difficult for many students. I think that there are several reasons for this. Writing well comes with practice, and most PhD students have had little experience with academic writing even if they have already completed an MSc. The same could be said of most early-career scientists (even after their PhDs), because practice is measured in completed pieces, it usually takes many years rather than weeks or months. You shouldn't expect that scientific writing will come automatically or immediately. In order to get this practice, you need to get started, and (again) many students find starting to write challenging. This guide was written to help you get started.

@@

In this part of the book, I start with an explanation of the hypothesis. As you will see, the hypothesis is central to all of the writing in each chapter or paper, and you need to have a very good idea of how to formulate your research question into a hypothesis before you begin. Then I build on helping you think through lots of different aspects of writing in science: style, citations, writing concisely, scientific names, and lots more. Finally, I introduce you to the idea of using a formula to structure your writing.

@@

When you read about the formula of writing a chapter (see [end of part 2](#formula)) you will see that everything in your chapter or paper revolves around the hypothesis or questions that you ask, normally at the end of your introduction (Figure \@ref(fig:Hypothesis)). In this section, we will take a look at how to write a hypothesis. This is a sticking point for many students. We are used to using and writing questions and statements in day-to-day communications, as well as reading popular media. But hypotheses (the plural of hypothesis) only rarely float across our desks. So how do we write one, and how do we know if our hypothesis is good?

In addition to this section, there is some good information out there on the web, and it's worth looking at this too: (e.g. [Wikihow](https://www.wikihow.com/Write-a-Hypothesis), [Wikipedia](https://en.wikipedia.org/wiki/Hypothesis), etc.). There's also some less good stuff out there, so read critically.

(ref:Hypothesis) **Generating a good hypothesis for your PhD chapter isn't easy.** A good hypothesis will be invaluable in helping you write the chapter. It is important to start working with hypotheses.

@@

A hypothesis is a statement of your research intent. It tells the reader (because just like all of your other written work, it is intended for an audience who will read it), what you plan to do in your research. But there's a little more to it than this. The hypothesis becomes a part of the scientific method if it is testable and (importantly) falsifiable, as well as being informed from previously published work on the subject.

Your hypothesis must be informed by the literature, which is why you spent so much time and effort crafting your introduction to inform your reader of the same. This is also why your hypothesis usually comes at the end of your introduction, because you spend all of the introduction telling your reader about it. There's not much point in writing more after the hypothesis, because once your reader has read that, they are ready to learn about how you went about testing it (in the [Materials & Methods](#methods)). The other important point to make is that the literature should dictate how you write your hypothesis, and the variables that you include. If, for example, you know that temperature is the most important variable, but all of the literature suggests that it is oxygen, you can't ignore oxygen and you should also frame your hypothesis using this variable (you can have more than one hypothesis after all!). In this case, you will also need to provide a sufficient introduction to temperature as a variable to justify its inclusion in your hypothesis. Perversely, your aim is not to prove that your idea is right, but to show that the hypothesis can be refuted.

We try to write a hypothesis that is falsifiable: i.e. you can prove (usually using statistical tests) that it is not correct (or at least show that the likelihood that it is correct is very low). That's why it is conventional to provide the 'null hypothesis' that is the falsified version of the statement, suggesting that there is no relationship between the variables you have proposed to measure. The convention is to label this 'null hypothesis' H~0~, while the 'alternative hypothesis' (the one that says your variables are related as you suggested) is written as H~1~. When you formulate your hypothesis, it is traditional to write your alternative hypothesis to indicate the directionality of your tested variables. This way, the reader can simply imply that the null hypothesis is when there is no relationship, but this will need to be stated if the null hypothesis is more complex.

@@

Most importantly, your hypothesis must come first, before you do the experiment or study. Hence the reason why this section comes at the start of part 2. Setting the hypothesis after the work is already done is fraudulent, and goes against the scientific method. Obviously, it isn't fair to pose the hypothesis once you already know the answer (also known as [HARKing](#TypeI)). This is why there is so much emphasis put on formulating your hypothesis during your research proposal. Getting it right will determine what you do and how you test it. If you think of an extra hypothesis that would be really useful to test once you've already done your study, you can conduct a post hoc test, but this should have more stringent levels of statistical assessment.

Writing a hypothesis isn't easy, but it is essential, and once you've understood what to do most of the rest of what you are writing for should make sense.

@@

Later in the book, I will emphasise that you must have introduced all the variables that you plan to use to test your hypothesis in your introduction. This usually comes in the second paragraph of your introduction, where you emphasise the utility of the dependent variable/s (what you are planning to measure) and your independent variable (what you will manipulate). Both of these variables should then feature in your hypothesis. Next, by paragraph four, you will have identified the problem that you are interested in tackling. In addition, your introduction will provide all of the pertinent literature that has relevance to this hypothesis, giving the all important context.

A simple way to consider making your hypothesis is to adopt an **"If. then. because."** construction where you add in your problem statement using your independent variable after 'if', and your prediction using your dependent variable after 'then', and finally the expected mechanism after 'because'. Using our example above with the "If. then. because." construction, we would say: "If environmental temperatures in which tadpoles develop are increased then tadpole development rate is faster because they follow the classic metabolism of ectotherms". Both independent variable (temperature) and dependent variable (tadpole development rate) are present in this hypothesis, and the predicted relationship between them is clear. In addition, the causal mechanism is stated. I say that this is a formulaic way to start writing your hypothesis, because it usually ends up as an inelegant statement, which can be better refined for a reader. A citation for your stated mechanism might also help clarify exactly where the justification for this comes from.

@@

Exogenous causal explanations concern mechanisms that are outside the body of individuals. Common exogenous mechanisms are climatic factors (temperature, humidity, precipitation, etc.) or may relate to the availability of food, predators or mates.

@@

### Teleological versus causal hypotheses

A teleological argument refers to the reason or a purpose of a particular process. For example, you may measure vertical migration of water fleas and suggest that diurnal migrations are made because the water fleas want to avoid predation. This is a teleological hypothesis because you are suggesting that the reason behind a process is the desire by water fleas to avoid predators. Although a reduction of predation may be a consequence of vertical migration in water fleas, each water flea does not think about predation and then starts it’s upward movement as a result. A common mistake made in biology is to apply teleological arguments to processes that have no purpose or reason. Evolution is often mistakenly suggested to have a purpose (e.g. to evolve to a more advanced state), but in fact, evolution is not a goal-orientated process. There is no end-point to evolution, and evolution did not start in order to meet some predetermined form or function. On the other hand, a causal hypothesis focuses on the factors about A that cause B.

### 2021-07-20 2.02-TypeI.Rmd
# Updating sections to chapters
Add files via upload
# After copy edits

## Being aware that you can get it wrong {#TypeI}

@@

With all the will in the world, when you are testing your hypothesis using statistics there is a chance that you will accept your alternative hypothesis when it is not valid. This is known as a ‘false positive’ or a Type I error (see Figure \@ref(fig:TypeITypeII)). It is also possible that you will accept your null hypothesis when you should have accepted your alternative hypothesis, also known as a Type II error, or a ‘false negative’. While it won't be any fun to get a Type II error, as scientists we should be more worried about Type I errors and the way in which they occur. This is because following a positive outcome, there is more chance that the work will be published, and that others may then pursue the same line of investigation mistakenly believing that their outcome is likely to be positive. Indeed, there are then lots of ways in which researchers may inadvertently or deliberately influence their outcomes towards Type I errors. This can even become a cultural bias that then permeates the literature.

Humans have a bias towards getting positive results [@trivers2011folly]. If you’ve put a lot of effort towards an experiment, then when you are interpreting your result you might feel motivated towards your reasoning making you more likely to accept your initial hypothesis. This is called [‘motivated reasoning’](https://youtu.be/w4RLfVxTGH4), and is a rational explanation why so many scientists get caught up in Type I errors. This is also known as a confirmation bias. Another manifestation of this is publication bias which also tends to be biased towards positive results, with as many as 84% of studies supporting their initial hypothesis [@fanelli2010positive]. Put together this means that scientists being human are more vulnerable to making Type I errors than Type II errors when evaluating their hypotheses. Although there are deliberate or inadvertent ways of making of Type I errors, here it is important to understand that simply by chance you can make a Type I error: accepting your alternative hypothesis even when it is not correct. To minimise this possibility to a reasonable level, most biologists set α (alpha) to 0.05 as an acceptable risk that they will encounter a false positive (Type I error).

(ref:TypeITypeII) **Type I and Type II errors.** In this figure, the columns refer to the truth regarding a hypothesis, even though the truth is unknown to the researcher. The rows are what the researcher finds when testing their hypothesis. The blue squares is what we are hoping to achieve when we set and test our hypothesis. The grey squares may happen if the hypothesis we set is indeed false. The other two possibilities are the false positive Type I error (red), and the false negative Type II error (black). In the table it seems that the chances of getting one of the four outcomes is equal, but in fact this is far from reality. There are several factors that can change the size of each potential outcome to testing a hypothesis.

@@

(ref:HighlyUnlikely) **The likelihood that highly unlikely hypotheses you test are accepted.** This figure has been redrawn after figure 1 in Forstmeier et al [-@forstmeier2017detecting]. This is a graphical representation of an argument first made by Ioannidis [-@ioannidis2005why].

@@

### Changing the likelihood of a hypothesis

The difference between Figure \@ref(fig:HighlyUnlikely) and Figures \@ref(fig:Unlikely) and \@ref(fig:Likely) is the likelihood that the hypotheses are correct changes. In the first one (Figure \@ref(fig:HighlyUnlikely)) we see highly unlikely hypotheses that will only be correct one in a hundred times. The blue squares denote the proportion of times in which the hypotheses are tested and found to be true. The black squares denote false negative findings, i.e a Type II error. The red squares denote false positive findings, i.e. a Type I error. Because the hypothesis is highly unlikely to be correct the majority of squares are light grey denoting that it was correctly found to be untrue. Because α is set at 0.05, we can expect that 5% of the 990 incorrect hypotheses will give us a false positive (49 or 50 times). Although it might seem unlikely that anyone would test such highly unlikely hypotheses, there are increasing numbers of governments around the world that create incentives for researchers to investigate what they term blue skies research, which might be better termed high risk research or investigations into highly unlikely hypotheses. The real problem with such hypotheses is that you are more likely to get a Type I error than actually find that your hypothesis is truly correct.

(ref:Unlikely) **The likelihood that unlikely hypotheses you test are accepted.** In this figure we see a scenario of unlikely hypotheses that are found to be correct approximately one in 10 times. Now we see that the possibility of committing a Type I error is roughly equivalent to a Type II error and to the probability of finding that the hypothesis is truly correct. Thus, if your result comes out positive, you are unlikely to know why.

@@

(ref:Likely) **The likelihood that likely hypotheses you test are accepted.** In this figure we see a scenario of likely hypotheses that are found to be correct approximately every other time.

@@

Now imagine some groundbreaking research (probably published in a high impact factor journal) that finds a highly unlikely (but presumably desirable) hypothesis is correct (our first scenario of 1 in 100). Researchers from other labs try to repeat the experiment, and some of them also find that they get positive results and get these published. None of the negative results get published, or the investigators don’t bother to submit them and move onto other areas of science. The positive results are written into text books and future generations of scientists test these types of hypotheses expecting them to be correct (our last scenario of 1 in 2). When the negative result is produced, a reasonable assumption is that a Type II error has been found, and that these scientists need to increase the power of their tests, inevitably increasing the cost of the experiment. But there are plenty of other ways that researchers could end up finding that their negative result is in fact positive, especially if this is what they are expecting. Simmons et al [-@simmons2011false] refer to this as the ‘researcher degrees of freedom’ in which researchers inadvertently or deliberately end up getting positive results because of the way in which they treat either their hypothesis or analyse their data. It is easy to see how a false scientific culture can be erected in these mistaken beliefs, and it is important to be aware of this possibility.

### Increasing the statistical power

In all of these figures the power of the analysis is set at 40%. The statistical power of any analysis depends on the sample size (or number of replicates) you're able to use. Some research has suggested that in most ecological and evolutionary studies this is actually more like 20% [see references in @forstmeier2017detecting]. What is important to notice in the next figure \@ref(fig:20PowerHyp) is that when we change the power of the analysis (in this case from 40% to 20%) we influence the proportion of Type II errors over finding that the hypothesis is correct. While the overall numbers of Type I errors does not change, if your analysis tells you to accept your alternative hypothesis, there is now a 1 in 5 chance (20%) that it will be a false positive (Type I error).

(ref:20PowerHyp) **The likelihood that likely hypotheses you test are accepted when you increase the power.** Here we see the scenario in which the hypotheses are quite likely to be correct one in two times. Now we can see that the possibility of creating a Type II error is highest. Next the blue squares show us the chances that we find the hypothesis is truly correct. Lastly, there's only 11% chance of a Type I error.

@@

What you should see when you look at these figures is that there is quite a high chance that we don't in fact correctly assign a true positive hypothesis. There's actually much more chance that we will commit a Type II error. Worse the more unlikely a hypothesis is, we are increasingly likely to commit the dreaded Type I error.

From the outset we should try to make sure that the hypotheses you are testing in your PhD are very likely to find positive results. In reality, this means that they are then iterations of hypotheses that are built on previous work. When you choose highly unlikely hypotheses, you need to be aware that this dramatically increases your chances of making a Type I error. The best way to overcome this is to look for multiple lines of evidence in your research. Once you have the most likely hypothesis that you can produce, you need to crank up your sampling so that you increase the power of your analysis, avoiding Type II errors.

In biological sciences we have more options than many scientific disciplines to investigate potentially false culture beliefs by using different model species, and looking at problems through different mechanisms. Presenting complimentary results from multiple lines of evidence within your thesis is an extremely powerful way to build evidence for your hypothesis. However, this requires creativity of experimental design as well as an open mind. It may mean that you have to draw upon the most fundamental of scientific abilities: to be comfortable with being wrong. It may also require dogged determination in terms of publishing negative results against the currently prevailing culture. As we shall see (in parts 4 and 5), publishing has another set of challenges.

>>>>

# Being aware that you can get it wrong {#TypeI}

@@

With all the will in the world, when you are testing your hypothesis using statistics, there is a chance that you will accept your alternative hypothesis when it is not valid. This is known as a ‘false positive’ or a Type I error (see Figure \@ref(fig:TypeITypeII)). It is also possible that you will accept your null hypothesis when you should have accepted your alternative hypothesis, also known as a Type II error, or a ‘false negative’. While it won't be any fun to get a Type II error, as scientists, we should be more worried about Type I errors and the way in which they occur. This is because following a positive outcome, there is more chance that the work will be published, and that others may then pursue the same line of investigation mistakenly believing that their outcome is likely to be positive. Indeed, there are then lots of ways in which researchers may inadvertently or deliberately influence their outcomes towards Type I errors. This can even become a cultural bias that then permeates the literature.

Humans have a bias towards getting positive results [@trivers2011folly]. If you’ve put a lot of effort towards an experiment, then when you are interpreting your result you might feel motivated towards your reasoning making you more likely to accept your initial hypothesis. This is called [‘motivated reasoning’](https://youtu.be/w4RLfVxTGH4), and is a rational explanation why so many scientists get caught up in Type I errors. This is also known as a confirmation bias. Another manifestation of this is publication bias which also tends to be biased towards positive results, with as many as 84% of studies supporting their initial hypothesis [@fanelli2010positive]. Thus scientists, being human, are more vulnerable to making Type I errors than Type II errors when evaluating their hypotheses. Although there are deliberate or inadvertent ways of making of Type I errors, here it is important to understand that simply by chance you can make a Type I error: accepting your alternative hypothesis even when it is not correct. To minimise this possibility to a reasonable level, most biologists set α (alpha) to 0.05 as an acceptable risk that they will encounter a false positive (Type I error).

(ref:TypeITypeII) **Type I and Type II errors.** In this figure, the columns refer to the truth regarding a hypothesis, even though the truth is unknown to the researcher. The rows are what the researcher finds when testing their hypothesis. The blue squares are what we are hoping to achieve when we set and test our hypothesis. The grey squares may happen if the hypothesis we set is indeed false. The other two possibilities are the false positive Type I error (red), and the false negative Type II error (black). In the figure it seems that the chances of getting one of the four outcomes is equal, but in fact, this is far from reality. There are several factors that can change the size of each potential outcome to testing a hypothesis.

@@

(ref:HighlyUnlikely) **The likelihood that highly unlikely hypotheses you test are accepted.** This figure has been redrawn after figure 1 in Forstmeier et al. [-@forstmeier2017detecting]. This is a graphical representation of an argument first made by Ioannidis [-@ioannidis2005why].

@@

## Changing the likelihood of a hypothesis

The difference between Figure \@ref(fig:HighlyUnlikely) and Figures \@ref(fig:Unlikely) and \@ref(fig:Likely) is the likelihood that the hypotheses are correct changes. In the first one (Figure \@ref(fig:HighlyUnlikely)) we see highly unlikely hypotheses that will only be correct one in a hundred times. The blue squares denote the proportion of times in which the hypotheses are tested and found to be true. The black squares denote false negative findings, i.e a Type II error. The red squares denote false positive findings, i.e. a Type I error. Because the hypothesis is highly unlikely to be correct the majority of squares are light grey denoting that it was correctly found to be untrue. Because α is set at 0.05, we can expect that 5% of the 990 incorrect hypotheses will give us a false positive (49 or 50 times). Although it might seem unlikely that anyone would test such highly unlikely hypotheses, there are increasing numbers of governments around the world that create incentives for researchers to investigate what they term 'blue skies' research, which might be better termed high-risk research or investigations into highly unlikely hypotheses. The real problem with such hypotheses is that you are more likely to get a Type I error than actually find that your hypothesis is truly correct.

(ref:Unlikely) **The likelihood that unlikely hypotheses you test are accepted.** In this figure, we see a scenario of unlikely hypotheses that are found to be correct approximately one in 10 times. Now we see that the possibility of committing a Type I error is roughly equivalent to a Type II error and to the probability of finding that the hypothesis is truly correct. Thus, if your result comes out positive, you are unlikely to know why.

@@

(ref:Likely) **The likelihood that likely hypotheses you test are accepted.** In this figure, we see a scenario of likely hypotheses that are found to be correct approximately every other time.

@@

Now imagine some groundbreaking research (probably published in a high impact factor journal) that finds a highly unlikely (but presumably desirable) hypothesis is correct (our first scenario of 1 in 100). Researchers from other labs try to repeat the experiment, and some of them also find that they get positive results and get these published. None of the negative results get published, or the investigators don’t bother to submit them and move onto other areas of science. The positive results are written into text books and future generations of scientists test these types of hypotheses expecting them to be correct (our last scenario of 1 in 2). When the negative result is produced, a reasonable assumption is that a Type II error has been found, and that these scientists need to increase the power of their tests, inevitably increasing the cost of the experiment. But there are plenty of other ways that researchers could end up finding that their negative result is, in fact, positive, especially if this is what they are expecting. Simmons et al. [-@simmons2011false] refer to this as the ‘researcher degrees of freedom’ in which researchers inadvertently or deliberately end up getting positive results because of the way in which they treat either their hypothesis or analyse their data. It is easy to see how a false scientific culture can be erected in these mistaken beliefs, and it is important to be aware of this possibility.

## Increasing the statistical power

In all of these figures, the power of the analysis is set at 40%. The statistical power of any analysis depends on the sample size (or number of replicates) you're able to use. Some research has suggested that in most ecological and evolutionary studies this is actually more like 20% [see references in @forstmeier2017detecting]. What is important to notice in the next figure (Figure \@ref(fig:20PowerHyp)) is that when we change the power of the analysis (in this case from 40% to 20%) we influence the proportion of Type II errors over finding that the hypothesis is correct. While the overall numbers of Type I errors do not change, if your analysis tells you to accept your alternative hypothesis, there is now a 1 in 5 chance (20%) that it will be a false positive (Type I error).

(ref:20PowerHyp) **The likelihood that likely hypotheses you test are accepted when you increase the power.** Here we see the scenario in which the hypotheses are quite likely to be correct one in two times. Now we can see that the possibility of creating a Type II error is highest. Next, the blue squares show us the chances that we find the hypothesis is truly correct. Lastly, there's only 11% chance of a Type I error.

@@

What you should see when you look at these figures is that there is quite a high chance that we don't correctly assign a true positive hypothesis. There's actually much more chance that we will commit a Type II error. Worse, the more unlikely a hypothesis is, we are increasingly likely to commit the dreaded Type I error.

From the outset, you should try to make sure that the hypotheses you are testing in your PhD are very likely to find positive results. In reality, this means that they are then iterations of hypotheses that are built on previous work. When you choose highly unlikely hypotheses, you need to be aware that this dramatically increases your chances of making a Type I error. The best way to overcome this is to look for multiple lines of evidence in your research. Once you have the most likely hypothesis that you can produce, you need to crank up your sampling so that you increase the power of your analysis, avoiding Type II errors.

In biological sciences, we have more options than many scientific disciplines to investigate potentially false cultural beliefs by using different model species, and looking at problems through different mechanisms. Presenting complementary results from multiple lines of evidence within your thesis is an extremely powerful way to build evidence for your hypothesis. However, this requires the creativity of experimental design as well as an open mind. It may mean that you have to draw upon the most fundamental of scientific abilities: to be comfortable with being wrong. It may also require dogged determination in terms of publishing negative results against the currently prevailing culture.

### 2021-07-20 2.03-NoHyp.Rmd
Add files via upload
# After copy edits

## What happens if you don’t have a hypothesis? {#nohypothesis}

So far we have concentrated on thesis chapters or manuscripts that have hypotheses to test. I have emphasised that there is a need to focus all of the sections of your writing to helping frame and understand the hypothesis (in the Introduction), explain your approach in collecting and testing the variables (in the Materials and Methods), provide the results of your tests (in the Results), and respond to the hypothesis (in the Discussion) together with placing it in context. But what, I hear you cry, if you don’t have a hypothesis to test?

@@

### Central problematic
If you have a particular problem that your research is trying to solve, this may result in a new methodology or perspective rather than testing a hypothesis. This work is still suitable for a journal manuscript (and there are specific journals dedicated to new methodologies). Instead of introducing your hypothesis, you can simply introduce the problem that your methodology applies to. A new methodology will probably best fill an existing gap, and so your introduction will likely point out what this gap is, provide evidence for why existing methodologies don’t fill the gap, and outline the variables in your novel approach.

@@

Hence, no matter what your central problem is, you still need to identify this and put it at the heart of your manuscript so that all of your sections address this point. Of course, there may be more than one, and if so you will need to clearly articulate each one and explain how they are related to each other. If you are unsure about how to move forward, the best thing to do is to look to the literature for other examples of what you’re trying to do. Someone has very likely done something similar before. Don’t forget to chat about it with your advisor, as they may have ideas about where to look.

### Does it matter that you don’t have a hypothesis?
Having a hypothesis makes your work stronger than if you only have a question or prediction. In each case, while your results might help answer your question or confirm your prediction, they won’t test the mechanism (so that you can’t explain why you have the results that you do), and they may not be falsifiable/refutable. Of course, Popper would argue that if you don’t have a null hypothesis, then your methodology isn’t scientific.

@@

### Avoid HARKing 
Hypothesising After Results Known (HARKing) is the practice of creating your hypothesis once the results are in and you’ve done some preliminary data analyses to see what is significant [@forstmeier2017detecting]. HARKing has become prevalent in science because of the confirmation bias by journals (only accepting papers that can statistically accept the alternative hypothesis), especially those with higher impact factor.

>>>>

# What happens if you don’t have a hypothesis? {#nohypothesis}

So far, we have concentrated on thesis chapters or manuscripts that have [hypotheses](#hypothesis) to test. I have emphasised that there is a need to focus all of the sections of your writing to helping frame and understand the hypothesis (in the [Introduction](#introduction)), explain your approach in collecting and testing the variables (in the [Materials and Methods](#methods)), provide the results of your tests (in the [Results](#results)), and respond to the hypothesis (in the [Discussion](#discussion)) together with placing it in context. But what, I hear you cry, if you don’t have a hypothesis to test?

@@

## Central problematic
If you have a particular problem that your research is trying to solve, this may result in a new methodology or perspective rather than testing a hypothesis. This work is still suitable for a journal manuscript (and there are specific journals dedicated to new methodologies). Instead of introducing your hypothesis, you can simply introduce the problem that your methodology applies to. A new methodology will probably best fill an existing gap, and so your introduction will likely point out what this gap is, provide evidence for why existing methodologies don’t fill the gap, and outline the variables in your novel approach.

@@

Hence, no matter what your central problem is, you still need to identify this and put it at the heart of your manuscript so that all of your sections address this point. Of course, there may be more than one, and if so, you will need to clearly articulate each one and explain how they are related to each other. If you are unsure about how to move forward, the best thing to do is to look to the literature for other examples of what you’re trying to do. Someone has very likely done something similar before. Don’t forget to chat about it with your advisor, as they may have ideas about where to look.

## Does it matter that you don’t have a hypothesis?
Having a hypothesis makes your work stronger than if you only have a question or prediction. In each case, while your results might help answer your question or confirm your prediction, they won’t test the mechanism (so that you can’t explain why you have the results that you do), and they may not be falsifiable/refutable. Of course, Popper would argue that if you don’t have a null hypothesis, then your methodology isn’t scientific.

@@

## Avoid HARKing 
Hypothesising After Results are Known (HARKing) is the practice of creating your hypothesis once the results are in and you’ve done some preliminary data analyses to see what is significant [@forstmeier2017detecting]. HARKing has become prevalent in science because of the confirmation bias by journals (only accepting papers that can statistically accept the alternative hypothesis), especially those with higher impact factor. Increasing chances of confirmation bias, or a [Type I error](#TypeI), is not desirable as you are more likely to accept the alternative hypothesis when it is not correct. Alternatively, you should seriously consider [preregistering your aims](#transparency) to remain transparent.

### 2021-07-20 2.04-BigIdea.Rmd
Add files via upload
# After copy edits

# What's the big idea? {#bigidea}
In the previous chapters, I've talked about the importance of having a hypothesis, and building that hypothesis in a logical framework within the introduction (see part 3). The introduction serves to inform the reader about why this particular hypothesis was chosen, introducing both the response and determinate variables, as well as the presumed mechanism by which the hypothesis can be falsified (or upheld).

## The funnel introduction 
But so much of what you read doesn't have any clear hypothesis, and there doesn't seem to be any thought about building on bigger concepts or ideas. In this chapter, I suggest that this might not be true only of papers on herpetological invasions (my specialist area), but also of herpetology in general (although I concede that some areas, such as herpetological physiology are actually quite good). Further, I contend that using these wider hypotheses or theories would actually be good for the authors concerned, as it would likely garner them a wider audience. Moreover, a greater number of biologists might come to realise how valuable reptiles and amphibians are as models in biology.

@@

## So where would we find these big ideas?
There are quite a few papers that synthesise hypotheses in various areas of biology. Here I provide a few to get you started. Ask your advisor about relevant reviews of hypotheses in your area of biological sciences. The first is by Mark Velland [-@vellend2010conceptual] on theories in community ecology. The next is by Jane Catford and colleagues [-@catford2009reducing] on hypotheses in invasion biology, but I encourage you to look for more up to date versions. Lastly, consider these wide ranging papers about so-called 'laws' in biology and ecology [@dhar2010laws; @lawton1999are]. Each of these papers will give you a list of big ideas, together with the citations for seminal papers that have built them. You will note that many of these theories are very old with some dating back to Darwin. You can think of the way that such ideas are structured as a hierarchy of hypotheses (Figure \@ref(fig:JeschkeHagar61)). Building such a hierarchy of ideas in your own specialist subject area is good reference material to prompt [critical reading](critical.html) and reflection. You may want to take this structure forward when thinking how to [introduce your PhD chapters](PhDreview.html).

(ref:JeschkeHagar61) **A hierarchy of hypothses can be used to help determine your Big Idea.** A useful structure for thinking about how hypotheses are structured was presented by @heger2018hierarchy in what they termed the _Hierarchy of Hypotheses_.

@@

**The take home message**

Reading the literature can really expand your mind and broaden your horizons. When undertaking a literature review [or when reviewing a paper], take the time to think about not only what has been tested, but what could have been. Make a list of theories and hypotheses in your own field and try to rank them in a hierarchy.

>>>>

# What's the big idea? {#bigidea}
In the previous chapters, I've talked about the importance of having a hypothesis, and building that hypothesis in a logical framework within the [introduction](#introduction). The introduction serves to inform the reader about why this particular hypothesis was chosen, introducing both the dependent and independent (response and determinate) variables, as well as the presumed mechanism by which the hypothesis can be falsified (or upheld; Figure \@ref(fig:Introfunnel)).

@@

## So where would we find these big ideas?
There are quite a few papers that synthesise hypotheses in various areas of biology. Here I provide a few to get you started. Ask your advisor about relevant reviews of hypotheses in your area of biological sciences. For example, here are some theories and hypotheses in ecology [@vellend2010conceptual; @travassos-britto2021pragmatic; @catford2009reducing], while these wide ranging papers about so-called 'laws' in biology and ecology [@dhar2010laws; @lawton1999are]. Each of these papers will give you a list of big ideas, together with the citations for seminal papers that have built them. Clearly, you will need to find the papers that set out these big ideas in your own discipline. You will note that many of the theories are very old, with some dating back to Darwin. You can think of the way that such ideas are structured as a hierarchy of hypotheses (Figure \@ref(fig:JeschkeHagar61)). Building such a hierarchy of ideas in your own specialist subject area is good reference material to prompt [critical reading](critical.html) and reflection. You may want to take this structure forward when thinking how to [introduce your PhD chapters](PhDreview.html).

(ref:JeschkeHagar61) **A hierarchy of hypotheses can be used to help determine your Big Idea.** A useful structure for thinking about how hypotheses are structured was presented by @heger2018hierarchy in what they termed the _Hierarchy of Hypotheses_.

@@

**The take-home message**

Reading the literature can really expand your mind and broaden your horizons. When undertaking a literature review, take the time to think about not only what has been tested, but what could have been. Make a list of theories and hypotheses in your own field and try to rank them in a hierarchy.
